How to Not write Virtual Machines for Dynamic Languages

### 2007-05-26 talk/dyla2007/dyla.tex
use proper quotes and long dashes

Dynamic languages are traditionally implemented by writing a virtual
machine for them, centered around an interpreter and/or a built-in
compiler and providing the object model and memory management. When
a language becomes more successful, the limitations of such an implementation
lead to the emergence of alternative implementations that try to solve some
of the problems. Another reason for new implementations is the desire to have
the language integrate well with existing, well-tuned object-oriented virtual
machine like Java's. In this paper, we describe the mechanisms that lead to an
abundance of implementations, and explore some of the limitations of standard
VMs.  We propose a different complementary alternative: flexibly generating
virtual machines instead
of writing them by hand.  This approach has been validated by the PyPy
project, with which we can automatically insert many features into the
generated virtual machines - including good just-in-time compilers tuned
to the dynamic language at hand.
\footnote{This research is partially supported by the EU funded %research
 project:
IST 004779 PyPy (PyPy: Implementing Python in Python).} \\

@@

For these reasons writing a virtual machine in C has many problems because it
forces the language implementer to deal with many low-level details. Limitations
of the C implementation lead to alternative implementations which draw
work-power from the reference implementation. An alternative to writing
implementations in C is to build them on top of one of the newer object oriented
virtual machines ("OO VM") such as the JVM or the CLR. This is often wanted by
the community anyway, since it leads to better re-usability of libraries of
these platforms. However, if a C implementation existed before the
implementation of such a VM is started, this enters in conflict with the goal of
having to maintain essentially a single, simple enough implementation for a
given programming language: as the language becomes popular, there will be a
demand for having it run on various platforms - high-level VMs as well as
C-level environments.

The argument we will make in the present paper is that it is possible to
benefit from and integrate with OO VMs while keeping the dynamic
language implemented by a single, simple source code base.  The idea is
to write an interpreter for that language in another sufficiently
high-level but less dynamic language.  This interpreter plays the role
of a specification for the dynamic language.  With a good enough
translation toolchain we can then generate whole virtual machines from
this specification - either full custom VMs for C-level operating
systems, or layers on top of various OO VMs.  In other words,
meta-programming techniques can be used to successfully replace a
foreseeable one-VM-fits-all standardization attempt.

@@

These limitations of a C-based implementation often lead to more
implementations of the language. For example in addition to the main C
implementation of Python ("CPython") there are a number of additional
implementations that want to fix some of the limitations of CPython. One of
them is "Stackless Python" which is a fork of CPython that adds micro-threading
capabilities to Python. One of the reasons for not incorporating it back into
CPython was that it was felt that they would make the implementation too
complex. Another implementation of the Python language is the Psyco project
which adds a JIT-compiler to CPython. As all other re-implementations, Psyco
contains the Python semantics and needs to be kept synchronized with CPython
manually.

@@

The present paper proposes to approach the implementation of dynamic
languages from a meta-level: virtual machines for such languages should
not be written by hand, but generated automatically "around" a
description of the language in the form of an interpreter for it.  We
argue that this approach gives many of the benefits usually expected by
an implementer when he decides to target an existing object-oriented
virtual machine.  It also gives other benefits that we will describe -
mostly in term of flexibility.  But most importantly, it lets a
community write a single source implementation of the language, avoiding
the time-consuming task of keeping multiple ones in sync.  The single
source can be used to generate either custom VMs for C-like
environments, or interpreters running on top of OO VMs.  It makes it
practical to experiment with large changes to the language and with
entirely new languages, like domain-specific languages, while at any
time being able to run the implemented language in a variety of
environments, from C/Posix to the JVM to .NET.

@@

\item
we use a very expressive \emph{object language} (RPython - an analyzable
subset of Python) as the language in which the complete Python
interpreter is written, together with the implementation of its
built-in types.  The language is still close to Python, e.g.  it is
object-oriented, provides rich built-in types and has automatic memory
management.  In other words, the source code of our complete Python
interpreter is mostly free of low-level details - no explicit memory
management, no pieces of C or C-level code.

@@

\item
this meta-programming component of PyPy is called the \emph{translation
framework}, as it translates RPython source code (i.e. the full Python
interpreter) into lower-level code.  Its purpose is to add aspects to
and specialize the interpreter to fit a selectable virtual or hardware
runtime environment.  This either turns the interpreter into a
standalone virtual machine, or integrates it into an existing OO VM.
The necessary support code - e.g. the garbage collector when
targeting C - is itself written in RPython in much the same spirit
that the Jikes RVM's GCs are written in Java \cite{JikesGC}; as needed, it is
translated together with the interpreter to form the final custom VM.
\end{itemize}

@@

Our approach - a single "meta-written" implementation - naturally leads
to language implementations that have various advantages over the
"hand-written" implementations.  First of all, it is a single-source
approach - we explicitly seek to solve the problem of proliferation of
implementations.  In the sequel we will show more precise evidence that
this can be done in a practical way with no major drawback.  By itself
this would already be a valid argument against the need for
standardization on a single OO VM.  But there are also other advantages
in generating language implementations. In our opinion these are very
significant, to the extent that it hints that meta-programming, though
not widely used in general-purpose programming, is an essential tool in
a language implementer's toolbox.

@@

A first point is that it makes interpreters easy to write, update and
generally experiment with.  More expressiveness helps at all levels: our
Python interpreter is written in RPython as a relatively simple
interpreter, in some respects easier to understand than CPython.  We are
using its high level and flexibility to quickly experiment with features
or implementation techniques in ways that would, in a traditional
approach, require pervasive changes to the source code.  For example,
PyPy's Python interpreter can optionally provide lazily computed objects
- a 150-lines extension in PyPy that would require global changes in
CPython.  Further examples can be found in our technical reports; we
should notably mention an extension adding a state-of-the-art security
model for Python based on data flow tacking \cite{D12.1}, and general
performance improvements found by extensive experimentation \cite{D06.1}, some
of which were back-ported to CPython.

If we compare with hand-writing an implementation for a specific OO VM,
then the latter requires not only good knowledge of the OO VM in
question and its object model - it requires the language implementer to
fit the language into the imposed models.  For example, it is natural to
map classes and instances of a dynamic object-oriented language to the
OO VM's notion of classes and instances, but this might not be a simple
task at all if the models are substantially different and/or if the OO
VM is essentially less dynamic than the language to implement.  In our
approach, this efforts is done at two levels: in a first step, while
writing the interpreter, the implementer does not need to worry about
integration with an OO VM's object model.  Of course, the integration
effort does not simply vanish - indeed, a simple translation of such an
interpreter to a given OO VM would give an interpreter for the dynamic
language which is unable to communicate with the host VM (which might
already be interesting in specific cases, but not in general).
Integration comes as a second step, and occurs at a different level, by
introducing mappings between the relevant classes of the interpreter and
the corresponding classes of the OO VM.  As of yet we have no evidence
that this makes the total integration effort much lower or higher; the
point is that it has proven possible to take the single, OO
VM-independent source code of PyPy's Python interpreter, and produce
from it a version that runs in and integrates with the rest of the .NET
environment - by writing \emph{orthogonal} code.

At the level of the translation framework, the ability to change or
insert new whole-program transformations makes some aspects of the
interpreter easier to deal with.  By "aspect" we mean, in the original
AOP sense, a feature that is added to an object program by a
meta-program.  The most obvious example in our context is the garbage
collector for the target environments that lack it.  The hand-written C
source of CPython is littered with macro calls that increment and
decrement reference counters.  Our translation framework can insert such
macro calls automatically - in fact, we have a choice of GCs, and
reference counting is only one of them (not a particularly efficient
one, either).  Some GCs require different transformations of the code.
By contrast, supporting more than one GC in CPython is close to
impossible without forking the whole source code base.

In fact, there is a well-known example of a CPython fork: Stackless
Python \cite{stackless}, which adds support for coroutines.  This is very hard to
do in CPython because the interpreter is written in a highly recursive
style.  Stackless Python required large-scale changes; it is not merged
back into CPython due to the pervasive increase in complexity that it
requires.  In PyPy, though, an optional "stackless" transformation is
able to turn the Python interpreter - also written in a simple highly
recursive style - into an efficient variant of continuation-passing
style (CPS), enabling the usage of coroutines in the translated
interpreter.  For more details and other examples of translation-level
transformations, see \cite{D07.1}.

\subsection{Getting good GCs and tools is possible}

XXX 
% ---- Bibliography ----
%\begin{small}
\bibliographystyle{abbrv}
\bibliography{dyla}
%\end{small}

>>>>

Dynamic languages are traditionally implemented by writing a virtual
machine for them, centered around an interpreter and/or a built-in
compiler and providing the object model and memory management. When
a language becomes more successful, the limitations of such an implementation
lead to the emergence of alternative implementations that try to solve some
of the problems. Another reason for new implementations is the desire to have
the language integrate well with existing, well-tuned object-oriented virtual
machine like Java's. In this paper, we describe the mechanisms that lead to an
abundance of implementations, and explore some of the limitations of standard
VMs.  We propose a different complementary alternative: flexibly generating
virtual machines instead
of writing them by hand.  This approach has been validated by the PyPy
project, with which we can automatically insert many features into the
generated virtual machines – including good just-in-time compilers tuned
to the dynamic language at hand.
\footnote{This research is partially supported by the EU funded %research
 project:
IST 004779 PyPy (PyPy: Implementing Python in Python).} \\

@@

For these reasons writing a virtual machine in C has many problems because it
forces the language implementer to deal with many low-level details. Limitations
of the C implementation lead to alternative implementations which draw
work-power from the reference implementation. An alternative to writing
implementations in C is to build them on top of one of the newer object oriented
virtual machines (``OO VM'') such as the JVM or the CLR. This is often wanted by
the community anyway, since it leads to better re-usability of libraries of
these platforms. However, if a C implementation existed before the
implementation of such a VM is started, this enters in conflict with the goal of
having to maintain essentially a single, simple enough implementation for a
given programming language: as the language becomes popular, there will be a
demand for having it run on various platforms – high-level VMs as well as
C-level environments.

The argument we will make in the present paper is that it is possible to
benefit from and integrate with OO VMs while keeping the dynamic
language implemented by a single, simple source code base.  The idea is
to write an interpreter for that language in another sufficiently
high-level but less dynamic language.  This interpreter plays the role
of a specification for the dynamic language.  With a good enough
translation toolchain we can then generate whole virtual machines from
this specification – either full custom VMs for C-level operating
systems, or layers on top of various OO VMs.  In other words,
meta-programming techniques can be used to successfully replace a
foreseeable one-VM-fits-all standardization attempt.

@@

These limitations of a C-based implementation often lead to more
implementations of the language. For example in addition to the main C
implementation of Python (``CPython'') there are a number of additional
implementations that want to fix some of the limitations of CPython. One of
them is ``Stackless Python'' which is a fork of CPython that adds micro-threading
capabilities to Python. One of the reasons for not incorporating it back into
CPython was that it was felt that they would make the implementation too
complex. Another implementation of the Python language is the Psyco project
which adds a JIT-compiler to CPython. As all other re-implementations, Psyco
contains the Python semantics and needs to be kept synchronized with CPython
manually.

@@

The present paper proposes to approach the implementation of dynamic
languages from a meta-level: virtual machines for such languages should
not be written by hand, but generated automatically ``around'' a
description of the language in the form of an interpreter for it.  We
argue that this approach gives many of the benefits usually expected by
an implementer when he decides to target an existing object-oriented
virtual machine.  It also gives other benefits that we will describe –
mostly in term of flexibility.  But most importantly, it lets a
community write a single source implementation of the language, avoiding
the time-consuming task of keeping multiple ones in sync.  The single
source can be used to generate either custom VMs for C-like
environments, or interpreters running on top of OO VMs.  It makes it
practical to experiment with large changes to the language and with
entirely new languages, like domain-specific languages, while at any
time being able to run the implemented language in a variety of
environments, from C/Posix to the JVM to .NET.

@@

\item
we use a very expressive \emph{object language} (RPython – an analyzable
subset of Python) as the language in which the complete Python
interpreter is written, together with the implementation of its
built-in types.  The language is still close to Python, e.g.  it is
object-oriented, provides rich built-in types and has automatic memory
management.  In other words, the source code of our complete Python
interpreter is mostly free of low-level details – no explicit memory
management, no pieces of C or C-level code.

@@

\item
this meta-programming component of PyPy is called the \emph{translation
framework}, as it translates RPython source code (i.e. the full Python
interpreter) into lower-level code.  Its purpose is to add aspects to
and specialize the interpreter to fit a selectable virtual or hardware
runtime environment.  This either turns the interpreter into a
standalone virtual machine, or integrates it into an existing OO VM.
The necessary support code – e.g. the garbage collector when
targeting C – is itself written in RPython in much the same spirit
that the Jikes RVM's GCs are written in Java \cite{JikesGC}; as needed, it is
translated together with the interpreter to form the final custom VM.
\end{itemize}

@@

Our approach – a single ``meta-written'' implementation – naturally leads
to language implementations that have various advantages over the
``hand-written'' implementations.  First of all, it is a single-source
approach – we explicitly seek to solve the problem of proliferation of
implementations.  In the sequel we will show more precise evidence that
this can be done in a practical way with no major drawback.  By itself
this would already be a valid argument against the need for
standardization on a single OO VM.  But there are also other advantages
in generating language implementations. In our opinion these are very
significant, to the extent that it hints that meta-programming, though
not widely used in general-purpose programming, is an essential tool in
a language implementer's toolbox.

@@

A first point is that it makes interpreters easy to write, update and
generally experiment with.  More expressiveness helps at all levels: our
Python interpreter is written in RPython as a relatively simple
interpreter, in some respects easier to understand than CPython.  We are
using its high level and flexibility to quickly experiment with features
or implementation techniques in ways that would, in a traditional
approach, require pervasive changes to the source code.  For example,
PyPy's Python interpreter can optionally provide lazily computed objects
– a 150-lines extension in PyPy that would require global changes in
CPython.  Further examples can be found in our technical reports; we
should notably mention an extension adding a state-of-the-art security
model for Python based on data flow tacking \cite{D12.1}, and general
performance improvements found by extensive experimentation \cite{D06.1}, some
of which were back-ported to CPython.

If we compare with hand-writing an implementation for a specific OO VM,
then the latter requires not only good knowledge of the OO VM in
question and its object model – it requires the language implementer to
fit the language into the imposed models.  For example, it is natural to
map classes and instances of a dynamic object-oriented language to the
OO VM's notion of classes and instances, but this might not be a simple
task at all if the models are substantially different and/or if the OO
VM is essentially less dynamic than the language to implement.  In our
approach, this efforts is done at two levels: in a first step, while
writing the interpreter, the implementer does not need to worry about
integration with an OO VM's object model.  Of course, the integration
effort does not simply vanish – indeed, a simple translation of such an
interpreter to a given OO VM would give an interpreter for the dynamic
language which is unable to communicate with the host VM (which might
already be interesting in specific cases, but not in general).
Integration comes as a second step, and occurs at a different level, by
introducing mappings between the relevant classes of the interpreter and
the corresponding classes of the OO VM.  As of yet we have no evidence
that this makes the total integration effort much lower or higher; the
point is that it has proven possible to take the single, OO
VM-independent source code of PyPy's Python interpreter, and produce
from it a version that runs in and integrates with the rest of the .NET
environment – by writing \emph{orthogonal} code.

At the level of the translation framework, the ability to change or
insert new whole-program transformations makes some aspects of the
interpreter easier to deal with.  By ``aspect'' we mean, in the original
AOP sense, a feature that is added to an object program by a
meta-program.  The most obvious example in our context is the garbage
collector for the target environments that lack it.  The hand-written C
source of CPython is littered with macro calls that increment and
decrement reference counters.  Our translation framework can insert such
macro calls automatically – in fact, we have a choice of GCs, and
reference counting is only one of them (not a particularly efficient
one, either).  Some GCs require different transformations of the code.
By contrast, supporting more than one GC in CPython is close to
impossible without forking the whole source code base.

In fact, there is a well-known example of a CPython fork: Stackless
Python \cite{stackless}, which adds support for coroutines.  This is very hard to
do in CPython because the interpreter is written in a highly recursive
style.  Stackless Python required large-scale changes; it is not merged
back into CPython due to the pervasive increase in complexity that it
requires.  In PyPy, though, an optional ``stackless'' transformation is
able to turn the Python interpreter – also written in a simple highly
recursive style – into an efficient variant of continuation-passing
style (CPS), enabling the usage of coroutines in the translated
interpreter.  For more details and other examples of translation-level
transformations, see \cite{D07.1}.

\subsection{Getting good GCs and tools is possible}
XXX

\section{Related Work}
XXX

\section{Conclusion}
XXX 
% ---- Bibliography ----
%\begin{small}
\bibliographystyle{abbrv}
\bibliography{dyla}
%\end{small}
